---
title: "Fhet_MDPP_MDPL_DE"
author: "Lisa Cohen"
date: "7/15/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# F. heteroclitus MDPL vs. MDPP: treatment is FW and transfer

## Which genes change in expression due to condition effect in each level of population (MDPP and MDPL)?

### Libraries
```{r echo=TRUE}
library(DESeq2)
library("lattice")
library(tximport)
library(readr)
library(gplots)
library(RColorBrewer)
source('~/Documents/UCDavis/ECE221/ECE221_final_project/extra_scripts/plotPCAWithSampleNames.R')
#setwd("~/Documents/Whitehead/Fhet_salmon_quant/combined_isoforms/")
```
### Import quantification counts from salmon (TPM)
```{r echo=TRUE}
dir<-"~/Documents/Whitehead/Fhet_salmon_quant"
files_list = list.files("~/Documents/Whitehead/Fhet_salmon_quant/combined_isoforms/")
files <- file.path(dir, "combined_isoforms",c("F_heteroclitusMDPL_FW_1.quant","F_heteroclitusMDPL_FW_2.quant","F_heteroclitusMDPL_FW_3.quant","F_heteroclitusMDPL_transfer_1.quant","F_heteroclitusMDPL_transfer_2.quant","F_heteroclitusMDPL_transfer_3.quant","F_heteroclitusMDPP_FW_1.quant","F_heteroclitusMDPP_FW_2.quant","F_heteroclitusMDPP_FW_3.quant","F_heteroclitusMDPP_transfer_1.quant","F_heteroclitusMDPP_transfer_2.quant","F_heteroclitusMDPP_transfer_3.quant"), "quant.sf")
#files <- files_list
names(files) <- c("MDPL_FW_1","MDPL_FW_2","MDPL_FW_3","MDPL_transfer_1","MDPL_transfer_2","MDPL_transfer_3","MDPP_FW_1","MDPP_FW_2","MDPP_FW_3","MDPP_transfer_1","MDPP_transfer_2","MDPP_transfer_3")
files
print(file.exists(files))
gene_names <- read.csv("~/Documents/Whitehead/Fhet_salmon_quant/Fhet_gene_transcript_names.csv")
cols<-c("row","transcript_id","gene_id")
colnames(gene_names)<-cols
tx2gene<-gene_names[,2:3]
head(tx2gene)
txi.salmon <- tximport(files, type = "salmon", tx2gene = tx2gene, reader = read_tsv)
head(txi.salmon$counts)
dim(txi.salmon$counts)
```
### Experimental Design

```{r echo=TRUE}
population = factor(c("MDPL","MDPL","MDPL","MDPL","MDPL","MDPL","MDPP","MDPP","MDPP","MDPP","MDPP","MDPP"))
condition = factor(c("FW","FW","FW","transfer","transfer","transfer","FW","FW","FW","transfer","transfer","transfer"))
ExpDesign <- data.frame(row.names=colnames(txi.salmon$counts), population = population,condition=condition)
ExpDesign
```

### DESeq2
```{r echo=TRUE}
dds <- DESeqDataSetFromTximport(txi.salmon, ExpDesign, ~population + condition + population:condition)
dds$population<-relevel(dds$population,ref="MDPP")
dds<-DESeq(dds,betaPrior=FALSE)
matrix(resultsNames(dds))
```

### Get normalized counts and filter out genes with low expression 
```{r echo=TRUE}
# get counts
counts_table = counts( dds, normalized=TRUE )
filtered_norm_counts<-counts_table[!rowSums(counts_table==0)>=1, ]
filtered_norm_counts<-as.data.frame(filtered_norm_counts)
GeneID<-rownames(filtered_norm_counts)
filtered_norm_counts<-cbind(filtered_norm_counts,GeneID)
dim(filtered_norm_counts)
head(filtered_norm_counts)
```
### Plot dispersion estimates
```{r echo=TRUE}
plotDispEsts(dds)
```

### PCA
```{r echo=TRUE}
log_dds<-rlog(dds)
plotPCAWithSampleNames(log_dds, intgroup="population", ntop=40000)
```

### Get GeneID and names from 
```{r echo=TRUE}
# get gene id names
gene_id<-read.csv("~/Documents/Whitehead/Fhet_salmon_quant/Fhet_gene_id_clean.csv")
cols<-c("row","GeneID","Name")
colnames(gene_id)<-cols
gene_id<-gene_id[,2:3]
dim(gene_id)
head(gene_id)
unique_genes<-unique(gene_id$GeneID)
length(unique_genes)
gene_id <- gene_id[unique(gene_id$GeneID),]
dim(gene_id)
head(gene_id)
```
### Get results table, merge with counts and add gene names
```{r echo=TRUE}
res<-results(dds, name="populationMDPL.conditiontransfer")
head(res)
res_ordered<-res[order(res$padj),]
GeneID<-rownames(res_ordered)
res_ordered<-as.data.frame(res_ordered)
res_genes<-cbind(res_ordered,GeneID)
dim(res_genes)
head(res_genes)
res_genes_merged <- merge(res_genes,filtered_norm_counts,by=unique("GeneID"))
dim(res_genes_merged)
head(res_genes_merged)
res_genes_merged_names <- merge(res_genes_merged,gene_id,by=unique("GeneID"))
dim(res_genes_merged_names)
head(res_genes_merged_names)
res_genes_merged_names <- res_genes_merged_names[order(res_genes_merged_names$padj),]
dim(res_genes_merged_names)
head(res_genes_merged_names)
##### removed duplicate GeneID
res_genes_merged_names_unique<-res_genes_merged_names[!duplicated(res_genes_merged_names$GeneID),]
res_ordered<-res_genes_merged_names_unique[order(res_genes_merged_names_unique$padj),]
head(res_ordered)
dim(res_ordered)
colnames(res_ordered)
res_ordered<-res_ordered[,c(1,20,2:19)]
write.csv(res_ordered, file="Fhet_MDPL_v_MDPP_condition_interactiononly_DESeq2_all.csv" )
```

### Get significantly differentially expressed genes, padj<0.05 and log2FC+=1
```{r echo=TRUE}
resSig = res_ordered[res_ordered$padj < 0.05, ]
resSig = resSig[resSig$log2FoldChange > 1 | resSig$log2FoldChange < -1,]
write.csv(resSig,file="Fhet_MDPL_v_MDPP_condition_interactiononly_DESeq_padj0.05_log2FC1.csv")
# MA plot with gene names
plot(log2(res_ordered$baseMean), res_ordered$log2FoldChange, col=ifelse(res_ordered$padj < 0.05, "red","gray67"),main="MDPL vs. MDPP (padj<0.05, log2FC = ±1)",xlim=c(1,20),pch=20,cex=1,ylim=c(-12,12))
abline(h=c(-1,1), col="blue")
```


### Labeled MA plot

```{r echo=TRUE}
# MA plot with gene names
# adjust significant genes to avoid crowded plotting
resSig = res_ordered[res_ordered$padj < 0.05, ]
resSig = resSig[resSig$log2FoldChange > 1 | resSig$log2FoldChange < -1,]
plot(log2(res_ordered$baseMean), res_ordered$log2FoldChange, col=ifelse(res_ordered$padj < 0.05, "red","gray67"),main="MDPL vs. transfer (padj<0.05, log2FC = ±1)",xlim=c(1,20),pch=20,cex=1,ylim=c(-12,12))
abline(h=c(-1,1), col="blue")
genes<-resSig$Name
mygenes <- resSig[,]
baseMean_mygenes <- mygenes[,"baseMean"]
log2FoldChange_mygenes <- mygenes[,"log2FoldChange"]
text(log2(baseMean_mygenes),log2FoldChange_mygenes,labels=genes,pos=2,cex=0.60)
```

### Heatmap

```{r echo=TRUE}
# heatmap
up_down<-resSig
dim(up_down)
#up_down_FC<-resSig
up_down_FC<-subset(up_down,up_down$log2FoldChange>2 | up_down$log2FoldChange< -2)
write.csv(up_down_FC,file="Fhet_DESeq_MDPL_v_MDPP_padj0.05_updownlogFC2.csv")
dim(up_down_FC)
d<-up_down_FC
d<-na.omit(d)
dim(d)
head(d)
colnames(d)
d<-up_down_FC[,c(9:20)]
d<-as.matrix(d)
d<-as.data.frame(d)
d<-as.matrix(d)
rownames(d) <- up_down_FC[,1]
#d<-type.convert(d)
head(d)
colnames(d)
#d<-d[,c(7,2,3,4,5,6,1,8)]
colnames(d)

hr <- hclust(as.dist(1-cor(t(d), method="pearson")), method="complete")
mycl <- cutree(hr, h=max(hr$height/1.5))
clusterCols <- rainbow(length(unique(mycl)))
myClusterSideBar <- clusterCols[mycl]
myheatcol <- greenred(75)
heatmap.2(d, main="MDPL vs. MDPP (padj<0.05, log2FC = ±2)", 
          Rowv=as.dendrogram(hr),
          cexRow=0.75,cexCol=0.8,srtCol= 90,
          adjCol = c(NA,0),offsetCol=2.5, 
          Colv=NA, dendrogram="row", 
          scale="row", col=myheatcol, 
          density.info="none", 
          trace="none", RowSideColors= myClusterSideBar)
```